\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-08-26 paper.qmd
Updated Figure 15 Caption

The MNIST dataset consists of grayscale images of handwritten digits [@lecun2010]. @yingfan2021 used this dataset to demonstrate how PaCMAP preserves non-linear structure in \pD{}. To evaluate whether PaCMAP provides a reasonable representation of the data, the \gD{} embedding of the handwritten digit 1 was selected. As shown in @fig-pacmap-author, the angle of the digit 1 images varies along the \gD{} structure.

@@

<!--PaCMAP param: n_components=2, n_neighbors=10, init=random, MN_ratio=0.9, FP_ratio=2.0-->
```{r}
#| fig-cap: "$2\\text{-}D$ layout from PaCMAP applied for the digit 1 of the MNIST dataset. The (hyper-)parameter settings, beyond the defaults, are 10 nearest neighbors, ratio of the number of mid-near pairs to the number of neighbors equal to 0.5, and the ratio of the number of further pairs to the number of neighbors is 2. We use our model-in-the-data-space to assess whether this is an accurate representation of structure present in the high-dimensional data, or if it is misleading. The angle of the digit 1 varies along this structure. Images at the top-left of the $2\\text{-}D$ layout show the digit 1 angled more to the left, while images at the bottom-right show the digit 1 angled more to the right."
#| label: fig-pacmap-author
#| fig-pos: H
#| out-height: 22%

@@

According to @fig-mnist1-sc1, the non-linear continuous structure observed in the \gD{} representation of PaCMAP (@fig-pacmap-author) is also visible when visualizing the model overlaid on the data space. This indicates that PaCMAP accurately captures the structure of the \pD{} data. Additionally, the model shows a twisted pattern within the non-linear structure in \pD{} space (@fig-mnist1-sc2), which is an additional pattern not visible in the \gD{} representation (@fig-pacmap-author). Furthermore, as shown in @fig-mnist1-sc3, some long edges exist in the \pD{} space that are not recognized as long edges in the \gD{} representation. However, PaCMAP is a reasonable \gD{} representation of MNIST digit 1 data, because PaCMAP preserves the non-linear structure present in the \pD{} data.

@@

There are certain data points that exhibit high error rates due to their deviation from the usual \pD{} data structure, which makes them anomalies (@fig-model-mnist b). These anomalies can be classified into two types: those that are anomalies within the non-linear structure and those that lie outside of it. The images associated with high model error points within the non-linear structure display different patterns of the digit 1, as shown in @fig-mnist1-within. However, when comparing these images to the ones found outside of the non-linear structure, it becomes evident that the latter display different patterns of the digit 1 (@fig-mnist1-out).

>>>>

The MNIST dataset consists of $70000$ grayscale images of handwritten digits [@lecun2010]. @yingfan2021 used this dataset to evaluate how PaCMAP preserves local structures in \gD{}. Because of the large number of images, we selected the handwritten digit 1 which has a non-linear structure in \gD{}. There are $7877$ images. Also, applied PCA as a preprocessing step and selected first $10$ principal components. The objective is to evaluate whether PaCMAP preserves the non-linear structure in \gD{}.

<!-- PaCMAP preserves non-linear structure in $784\text{-}D$. To evaluate whether PaCMAP provides a reasonable representation of the data, the \gD{} embedding of the handwritten digit 1 was selected. As shown in @fig-pacmap-author, the angle of the digit 1 images varies along the \gD{} structure. -->

@@

<!--PaCMAP param: n_components=2, n_neighbors=10, init=random, MN_ratio=0.9, FP_ratio=2.0-->
```{r}
#| fig-cap: "$2\\text{-}D$ layout from PaCMAP applied for the digit 1 of the MNIST dataset. The (hyper-)parameter settings, beyond the defaults, are $10$ nearest neighbors, ratio of the number of mid-near pairs to the number of neighbors equal to $0.5$, and the ratio of the number of further pairs to the number of neighbors is $2$. We use our *model-in-the-data-space* to assess whether this is an accurate representation of structure present in $10\\text{-}D$, or if it is misleading. The angle of the digit 1 varies along this structure. Images at the top-left of the $2\\text{-}D$ layout show the digit 1 angled more to the left, while images at the bottom-right show the digit 1 angled more to the right."
#| label: fig-pacmap-author
#| fig-pos: H
#| out-height: 22%

@@

According to @fig-mnist1-sc1, the non-linear continuous structure observed in the \gD{} representation of PaCMAP (@fig-pacmap-author) is also visible when visualizing the model overlaid on the data space. This indicates that PaCMAP accurately captures the structure of the $10\text{-}D$ data. Additionally, the model shows a twisted pattern within the non-linear structure in $10\text{-}D$ space (@fig-mnist1-sc2), which is an additional pattern not visible in the \gD{} representation (@fig-pacmap-author). Furthermore, as shown in @fig-mnist1-sc3, some long edges exist in the $10\text{-}D$ space that are not recognized as long edges in the \gD{} representation. However, PaCMAP is a reasonable \gD{} representation of MNIST digit 1 data, because PaCMAP preserves the non-linear structure present in the $10\text{-}D$ data.

@@

There are certain data points that exhibit high error rates due to their deviation from the usual $10\text{-}D$ data structure, which makes them anomalies (@fig-model-mnist b). These anomalies can be classified into two types: those that are anomalies within the non-linear structure and those that lie outside of it. The images associated with high model error points within the non-linear structure display different patterns of the digit 1, as shown in @fig-mnist1-within. However, when comparing these images to the ones found outside of the non-linear structure, it becomes evident that the latter display different patterns of the digit 1 (@fig-mnist1-out).

### 2024-08-26 paper.tex
Updated UMAP proj
Updated Figure 15 Caption

The MNIST dataset consists of grayscale images of handwritten digits
\citep{lecun2010}. \citet{yingfan2021} used this dataset to demonstrate
how PaCMAP preserves non-linear structure in \pD{}. To evaluate whether
PaCMAP provides a reasonable representation of the data, the \gD{}
embedding of the handwritten digit 1 was selected. As shown in
Figure~\ref{fig-pacmap-author}, the angle of the digit 1 images varies
along the \gD{} structure.

@@

\caption{\label{fig-pacmap-author}\(2\text{-}D\) layout from PaCMAP
applied for the digit 1 of the MNIST dataset. The (hyper-)parameter
settings, beyond the defaults, are 10 nearest neighbors, ratio of the
number of mid-near pairs to the number of neighbors equal to 0.5, and
the ratio of the number of further pairs to the number of neighbors is
2. We use our model-in-the-data-space to assess whether this is an
accurate representation of structure present in the high-dimensional
data, or if it is misleading. The angle of the digit 1 varies along this
structure. Images at the top-left of the \(2\text{-}D\) layout show the
digit 1 angled more to the left, while images at the bottom-right show
the digit 1 angled more to the right.}

@@

According to \textbf{?@fig-mnist1-sc1}, the non-linear continuous
structure observed in the \gD{} representation of PaCMAP
(Figure~\ref{fig-pacmap-author}) is also visible when visualizing the
model overlaid on the data space. This indicates that PaCMAP accurately
captures the structure of the \pD{} data. Additionally, the model shows
a twisted pattern within the non-linear structure in \pD{} space
(\textbf{?@fig-mnist1-sc2}), which is an additional pattern not visible
in the \gD{} representation (Figure~\ref{fig-pacmap-author}).
Furthermore, as shown in \textbf{?@fig-mnist1-sc3}, some long edges
exist in the \pD{} space that are not recognized as long edges in the
\gD{} representation. However, PaCMAP is a reasonable \gD{}
representation of MNIST digit 1 data, because PaCMAP preserves the
non-linear structure present in the \pD{} data.

@@

There are certain data points that exhibit high error rates due to their
deviation from the usual \pD{} data structure, which makes them
anomalies (Figure~\ref{fig-model-mnist} b). These anomalies can be
classified into two types: those that are anomalies within the
non-linear structure and those that lie outside of it. The images
associated with high model error points within the non-linear structure
display different patterns of the digit 1, as shown in
Figure~\ref{fig-mnist1-within}. However, when comparing these images to
the ones found outside of the non-linear structure, it becomes evident
that the latter display different patterns of the digit 1
(Figure~\ref{fig-mnist1-out}).

>>>>

The MNIST dataset consists of \(70000\) grayscale images of handwritten
digits \citep{lecun2010}. \citet{yingfan2021} used this dataset to
evaluate how PaCMAP preserves local structures in \gD{}. Because of the
large number of images, we selected the handwritten digit 1 which has a
non-linear structure in \gD{}. There are \(7877\) images. Also, applied
PCA as a preprocessing step and selected first \(10\) principal
components. The objective is to evaluate whether PaCMAP preserves the
non-linear structure in \gD{}.

@@

\caption{\label{fig-pacmap-author}\(2\text{-}D\) layout from PaCMAP
applied for the digit 1 of the MNIST dataset. The (hyper-)parameter
settings, beyond the defaults, are \(10\) nearest neighbors, ratio of
the number of mid-near pairs to the number of neighbors equal to
\(0.5\), and the ratio of the number of further pairs to the number of
neighbors is \(2\). We use our \emph{model-in-the-data-space} to assess
whether this is an accurate representation of structure present in
\(10\text{-}D\), or if it is misleading. The angle of the digit 1 varies
along this structure. Images at the top-left of the \(2\text{-}D\)
layout show the digit 1 angled more to the left, while images at the
bottom-right show the digit 1 angled more to the right.}

@@

According to \textbf{?@fig-mnist1-sc1}, the non-linear continuous
structure observed in the \gD{} representation of PaCMAP
(Figure~\ref{fig-pacmap-author}) is also visible when visualizing the
model overlaid on the data space. This indicates that PaCMAP accurately
captures the structure of the \(10\text{-}D\) data. Additionally, the
model shows a twisted pattern within the non-linear structure in
\(10\text{-}D\) space (\textbf{?@fig-mnist1-sc2}), which is an
additional pattern not visible in the \gD{} representation
(Figure~\ref{fig-pacmap-author}). Furthermore, as shown in
\textbf{?@fig-mnist1-sc3}, some long edges exist in the \(10\text{-}D\)
space that are not recognized as long edges in the \gD{} representation.
However, PaCMAP is a reasonable \gD{} representation of MNIST digit 1
data, because PaCMAP preserves the non-linear structure present in the
\(10\text{-}D\) data.

@@

There are certain data points that exhibit high error rates due to their
deviation from the usual \(10\text{-}D\) data structure, which makes
them anomalies (Figure~\ref{fig-model-mnist} b). These anomalies can be
classified into two types: those that are anomalies within the
non-linear structure and those that lie outside of it. The images
associated with high model error points within the non-linear structure
display different patterns of the digit 1, as shown in
Figure~\ref{fig-mnist1-within}. However, when comparing these images to
the ones found outside of the non-linear structure, it becomes evident
that the latter display different patterns of the digit 1
(Figure~\ref{fig-mnist1-out}).
